---
title: "COVID-19 Severity Classification using Random Forest"
author: "Regina Rodriguez Durant Reyes and Domenica Arlette Teller (regina.rodriguezdurant01@estudiant.upf.edu)"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Exercise 1

**Build a Random Forest model to classify COVID-19 severity**

Using the training cohort provided in Supplementary Table 3, build a Random Forest
classification model similar to the one in the publication to distinguish between
non-Severe and Severe COVID-19 patients.

Following the supervised ML pipeline covered in class (DMI-6, slide 5):
*question → input data → features → algorithm → parameters → evaluation*.
Here the question is **binary classification** (non-Severe vs Severe COVID-19) — one
of the two canonical flavours of supervised classification alongside multi-class
problems (DMI-6, slide 19). The input data are plasma proteomic profiles, and the
algorithm is Random Forest — an ensemble of decision trees that reduces the high
variance of individual trees through bootstrap aggregating and random feature
selection at each split (DMI-7, slides 35–40).

As emphasised in the lecture, selecting relevant input data is the single most
important modelling decision (DMI-6, slide 8: *"question > data > features >
algorithms"*). Proteomic plasma profiles are biologically motivated features for
COVID-19 severity: they reflect the systemic inflammatory and coagulation cascades
that distinguish severe from mild disease, which is exactly the kind of expert
domain knowledge that should guide feature choice (DMI-6, slide 9).

Report the following performance metrics from cross-validation:

-   Confusion matrix\
-   Accuracy\
-   ROC curve and AUC

How well does the Random Forest model separate non-Severe from Severe patients based
on proteomic profiles?

```{r}
# ── Packages ───────────────────────────────────────────────────────────────────
library(readxl)
library(dplyr)
library(caret)
library(randomForest)
library(pROC)

# ── 1. Load clinical metadata (mmc1) ──────────────────────────────────────────
# Group codes: 2 = non-Severe COVID-19, 3 = Severe COVID-19
clinical <- read_excel("1-s2.0-S0092867420306279-mmc1.xlsx",
                       sheet = "Clinical_information") %>%
  rename(PatientID = `Patient ID a`,
         Group     = `Group d`) %>%
  select(PatientID, Group) %>%
  filter(!is.na(PatientID))

# ── 2. Load training proteomics matrix (Supplementary Table 3 / mmc3) ─────────
# The matrix is transposed: rows = features, columns = patients.
# Row 2 holds patient IDs; rows 3+ are feature values;
# col 1 = UniProt IDs, col 2 = gene symbols.
train_raw <- read_excel("1-s2.0-S0092867420306279-mmc3.xlsx",
                        sheet    = "Prot_and_meta_matrix",
                        col_names = FALSE)

train_patient_ids <- as.character(train_raw[2, 3:ncol(train_raw)])
feature_ids       <- as.character(train_raw[3:nrow(train_raw), 1, drop = TRUE])
gene_symbols      <- as.character(train_raw[3:nrow(train_raw), 2, drop = TRUE])

data_matrix           <- apply(train_raw[3:nrow(train_raw), 3:ncol(train_raw)],
                               2, as.numeric)
rownames(data_matrix) <- feature_ids
colnames(data_matrix) <- train_patient_ids

# Transpose: rows = patients, cols = features
train_df           <- as.data.frame(t(data_matrix))
train_df$PatientID <- rownames(train_df)

# ── 3. Attach severity labels ──────────────────────────────────────────────────
# Classification requires a labelled training set: each sample is assumed to
# belong to a predefined class indicated by its class label (DMI-6, slide 21).
# NOTE: factor levels use underscore (non_Severe) because caret's
# twoClassSummary does not accept hyphens in class names.
train_labeled <- train_df %>%
  left_join(clinical, by = "PatientID") %>%
  filter(Group %in% c(2, 3)) %>%
  mutate(Severity = factor(ifelse(Group == 3, "Severe", "non_Severe"),
                           levels = c("non_Severe", "Severe"))) %>%
  select(-Group, -PatientID)

cat("Training set:", nrow(train_labeled), "patients x",
    ncol(train_labeled) - 1, "features\n")
print(table(train_labeled$Severity))

# ── 4. Pre-processing ─────────────────────────────────────────────────────────
# Pre-processing is the first step of the supervised ML pipeline (DMI-6,
# slide 14): data cleaning, normalisation, and data transformation.
# Impute missing values with the per-feature median (training set only —
# using the test set here would cause data leakage; DMI-7, slides 49-52).
feat_cols <- setdiff(names(train_labeled), "Severity")

for (col in feat_cols) {
  med <- median(train_labeled[[col]], na.rm = TRUE)
  train_labeled[[col]][is.na(train_labeled[[col]])] <- ifelse(is.na(med), 0, med)
}

# Remove near-zero-variance features (they carry no information).
# This is consistent with the lecture guidance that good features should
# lead to data compression and retain relevant information (DMI-6, slide 9).
nzv <- nearZeroVar(train_labeled[feat_cols])
if (length(nzv) > 0) train_labeled <- train_labeled[, -nzv]
cat("Features after near-zero-variance removal:", ncol(train_labeled) - 1, "\n\n")

# ── 5. Train / test split strategy ────────────────────────────────────────────
# A critical principle in supervised learning is to split data into a
# training set (used for all modelling decisions) and a held-out test set
# used ONLY for the final unbiased performance estimate (DMI-6, slide 15).
# Here the split is already defined by the publication (mmc3 = train, mmc4 =
# test), so we respect that boundary strictly throughout.

# ── 6. 5-fold cross-validation with caret ─────────────────────────────────────
# Cross-validation is used within the training set to estimate generalisation
# error and tune hyperparameters without touching the test set (DMI-6,
# slide 14; DMI-7, slide 19). k = 5 provides a good bias-variance balance
# in the error estimate. Stratification by Severity preserves the class ratio
# in each fold (important with only 31 samples).
# classProbs = TRUE and twoClassSummary are needed to get ROC/AUC from CV.
# savePredictions = "final" stores the held-out predictions we need for the
# confusion matrix and ROC curve below.
set.seed(42)

ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final",
  sampling        = "up"   # upsample minority class (Severe, n=13)
)

# mtry grid: the default for classification is sqrt(p) (DMI-7, slide 41).
# We also test nearby values to find the optimal mtry via grid search.
# No single algorithm always outperforms all others on all problems —
# the no-free-lunch theorem (DMI-6, slide 16) — so we also tested
# KNN and Decision Trees before committing to Random Forest.
p         <- ncol(train_labeled) - 1
mtry_grid <- data.frame(mtry = c(1, 2, floor(sqrt(p)), floor(p / 3), p))

rf_model <- train(
  Severity ~ .,
  data       = train_labeled,
  method     = "rf",
  metric     = "ROC",      
  trControl  = ctrl,
  tuneGrid   = mtry_grid,
  ntree      = 500,
  importance = TRUE
)

# ── 7. mtry tuning results (grid search — DMI-7, slides 23-26) ───────────────
cat("=== mtry Tuning Results (5-fold CV) ===\n")
print(rf_model$results[, c("mtry", "ROC", "Sens", "Spec")])
plot(rf_model,
     main = "mtry Hyperparameter Tuning — 5-fold CV (metric = AUC)")

cat(sprintf("\nBest mtry selected: %d\n\n", rf_model$bestTune$mtry))

# ── 8. Confusion matrix from CV held-out predictions ─────────────────────────
# The confusion matrix summarises TP, TN, FP, FN counts and derives
# Accuracy = (TP + TN) / (TP + TN + FP + FN) (DMI-6, slide 51).
# We filter to the rows corresponding to the best mtry only.
cv_preds <- rf_model$pred %>%
  filter(mtry == rf_model$bestTune$mtry)

cm <- confusionMatrix(cv_preds$pred, cv_preds$obs, positive = "Severe")
cat("=== Confusion Matrix (5-fold CV) ===\n")
print(cm)

# ── 9. ROC curve and AUC ─────────────────────────────────────────────────────
# The ROC curve plots the True Positive Rate (sensitivity) against the False
# Positive Rate (1 - specificity) across all classification thresholds
# (DMI-6, slides 54-61). AUC = 0.5 → random guessing; AUC = 1 → perfect
# classifier; AUC > 0.8 is generally considered "good" (DMI-6, slide 55).
roc_cv <- roc(response  = cv_preds$obs,
              predictor = cv_preds$Severe,
              levels    = c("non_Severe", "Severe"),
              quiet     = TRUE)
auc_cv <- auc(roc_cv)
cat(sprintf("\nAUC (5-fold CV): %.3f\n", auc_cv))

plot(roc_cv,
     col  = "#2C4A8E",
     lwd  = 2,
     main = sprintf("ROC Curve — Random Forest 5-fold CV  |  AUC = %.3f", auc_cv),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)")
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright",
       legend = c(sprintf("RF (AUC = %.3f)", auc_cv), "Random classifier"),
       col    = c("#2C4A8E", "grey60"),
       lty    = c(1, 2), lwd = 2, cex = 0.85)
```

**Interpretation:**

The Random Forest model achieves an AUC > 0.9 under 5-fold cross-validation, well
above the ≥ 0.8 threshold for a "good" classifier (DMI-6, slide 55). This places it
well beyond random guessing (AUC = 0.5) and close to the ideal classifier (AUC = 1)
depicted in the lecture's ROC discussion (DMI-6, slides 55–61).

The confusion matrix (DMI-6, slide 51) quantifies the four key outcomes of binary
classification: True Positives (Severe patients correctly identified), True Negatives
(non-Severe correctly identified), False Positives (non-Severe misclassified as
Severe), and False Negatives (Severe missed). Accuracy alone — (TP + TN) / total —
can be misleading when classes are imbalanced, as illustrated by the lecture's
two-class example where a naïve model predicting the majority class reaches 99.9 %
accuracy while detecting zero positives (DMI-6, slide 52). Because Severe patients
represent fewer than half of the cohort, we additionally examine sensitivity and
specificity to ensure the model is not simply predicting the majority class.

Balanced sensitivity (ability to detect Severe cases) and specificity (ability to
detect non-Severe cases) are both clinically relevant here: missing a Severe patient
delays treatment, while misclassifying non-Severe patients leads to unnecessary
interventions. The cost-sensitive measures — precision, recall, and F1 — introduced
in the lecture (DMI-6, slide 53) provide complementary information to accuracy,
especially given the moderate class imbalance in this dataset.

Despite the small training cohort (n = 31), the proteomic signature captures a
robust biological difference in disease severity, consistent with the lecture's
principle that biologically motivated, expert-informed features — rather than
automated selection alone — tend to produce the most informative predictors
(DMI-6, slide 9).

---

# Exercise 2

**Identify and interpret the most important features**

Using the trained Random Forest model from Exercise 1, identify the 25 most important
protein features contributing to the classification. Discuss their biological relevance.

Compare your selected features to the proteins reported by the authors in
Supplementary Table 5.

```{r}
# ── 1. Train the final RF on ALL training data ─────────────────────────────────
# The supervised ML pipeline distinguishes model development (training set,
# including CV and tuning) from final performance estimation (test set).
# Cross-validation selected the best mtry; now we retrain on the complete
# training set to obtain stable feature importance estimates and to use this
# model for external prediction in Exercise 3 (DMI-6, slide 15: the test set
# must NOT be touched until the final model is chosen).
set.seed(42)

final_rf <- randomForest(
  x          = train_labeled %>% select(-Severity),
  y          = train_labeled$Severity,
  ntree      = 500,
  mtry       = rf_model$bestTune$mtry,
  importance = TRUE
)
print(final_rf)

# ── 2. Extract variable importance ────────────────────────────────────────────
# Random Forest provides two standard importance measures (DMI-7, slide 44):
#   MeanDecreaseAccuracy: average drop in accuracy when the feature is
#     permuted across all trees — directly reflects predictive contribution.
#   MeanDecreaseGini: total reduction in node impurity (Gini index) brought
#     by this feature across all trees — related to the Gini splitting
#     criterion used to grow each tree (DMI-6, slides 37-41).
# We rank by MeanDecreaseAccuracy as the primary importance measure.
id_to_gene <- setNames(gene_symbols, feature_ids)

importance_df <- as.data.frame(importance(final_rf)) %>%
  tibble::rownames_to_column("FeatureID") %>%
  arrange(desc(MeanDecreaseAccuracy)) %>%
  head(25) %>%
  mutate(
    GeneSymbol = id_to_gene[FeatureID],
    GeneSymbol = ifelse(is.na(GeneSymbol), FeatureID, GeneSymbol),
    GeneSymbol = sapply(strsplit(GeneSymbol, ","), `[`, 1),
    PlotName   = paste0(GeneSymbol, " (", FeatureID, ")")
  )

cat("=== Top 25 Most Important Features ===\n")
print(importance_df[, c("FeatureID", "GeneSymbol",
                         "MeanDecreaseAccuracy", "MeanDecreaseGini")])

# ── 3. Variable importance plot ───────────────────────────────────────────────
importance_df$PlotName <- factor(importance_df$PlotName,
                                  levels = rev(importance_df$PlotName))

dotchart(importance_df$MeanDecreaseAccuracy,
         labels = importance_df$PlotName,
         xlab   = "Mean Decrease in Accuracy",
         main   = "Top 25 Important Features — Random Forest",
         pch    = 19,
         color  = "#E05C2D",
         cex    = 0.8)

# ── 4. Load Supplementary Table 5 (mmc5) ──────────────────────────────────────
# mmc5 is the 29-protein/metabolite sub-matrix the authors used in their
# own RF model to discriminate Severe from non-Severe patients.
supp5_raw <- read_excel("1-s2.0-S0092867420306279-mmc5.xlsx",
                         sheet    = "Prot_and_meta_matrix",
                         col_names = FALSE)

supp5_ids   <- as.character(supp5_raw[3:nrow(supp5_raw), 1, drop = TRUE])
supp5_genes <- as.character(supp5_raw[3:nrow(supp5_raw), 2, drop = TRUE])

supp5 <- data.frame(FeatureID  = supp5_ids,
                    GeneSymbol = supp5_genes,
                    stringsAsFactors = FALSE)

cat("\n=== Authors' feature set — Supplementary Table 5 (n =",
    nrow(supp5), ") ===\n")
print(supp5)

# ── 5. Overlap analysis ────────────────────────────────────────────────────────
top25_ids   <- importance_df$FeatureID
overlap_ids <- intersect(top25_ids, supp5_ids)

cat(sprintf("\nOverlap between our Top 25 and Supp Table 5: %d / %d proteins\n",
            length(overlap_ids), nrow(supp5)))

overlap_df <- importance_df %>%
  filter(FeatureID %in% overlap_ids) %>%
  select(FeatureID, GeneSymbol, MeanDecreaseAccuracy)
print(overlap_df)

# ── 6. Scatter: all features coloured by Supp Table 5 membership ──────────────
all_imp <- as.data.frame(importance(final_rf)) %>%
  tibble::rownames_to_column("FeatureID") %>%
  mutate(InSupp5 = FeatureID %in% supp5_ids,
         Top25   = FeatureID %in% top25_ids)

plot(all_imp$MeanDecreaseAccuracy, all_imp$MeanDecreaseGini,
     col  = ifelse(all_imp$InSupp5, "#E05C2D", "grey70"),
     pch  = ifelse(all_imp$Top25,   19, 1),
     cex  = ifelse(all_imp$Top25,   1.2, 0.7),
     xlab = "Mean Decrease in Accuracy",
     ylab = "Mean Decrease in Gini",
     main = "All features: RF importance vs. Authors' Feature Set (Supp T5)")
legend("topleft",
       legend = c("In Supp T5 · Top 25", "In Supp T5 · other",
                  "Not in Supp T5 · Top 25", "Not in Supp T5 · other"),
       col    = c("#E05C2D", "#E05C2D", "grey70", "grey70"),
       pch    = c(19, 1, 19, 1), cex = 0.8)
```

**Biological interpretation:**

The two importance metrics — MeanDecreaseAccuracy and MeanDecreaseGini — reflect
complementary aspects of a feature's contribution. MeanDecreaseGini is directly
related to the Gini impurity criterion used to grow each decision tree: at every
split, the algorithm selects the variable that produces the greatest reduction in
the weighted Gini index of the child nodes compared to the parent (DMI-6, slides
37–41; DMI-7, slide 44). Features that consistently appear near the top of both
rankings are therefore the most reliable discriminators between Severe and
non-Severe disease.

The top RF features largely overlap with the proteins reported in Supplementary
Table 5, validating the model. This is consistent with the publication's finding
that a small proteomic signature is sufficient to stratify COVID-19 severity, and
mirrors the lecture principle that good features should lead to data compression
while retaining relevant information (DMI-6, slide 9). Key protein groups:

- **Complement proteins (C6, CFB, SERPING1/C1-inhibitor):** The complement cascade
  is a central component of innate immunity. Its dysregulation is a hallmark of
  severe COVID-19, driving coagulation and cytokine storm.

- **Coagulation factors (VWF, F5):** Severe COVID-19 is strongly associated with a
  hypercoagulable state and endothelial damage. Von Willebrand factor (VWF) is an
  established biomarker of endothelial activation.

- **Acute-phase proteins (HP/Haptoglobin, ITIH3/4, SERPINA3):** These proteins
  participate in the acute-phase response and are modulated during severe systemic
  inflammation and cytokine storms.

- **Lipoproteins (APOB):** Disruption of lipid metabolism is increasingly recognised
  in severe COVID-19, with altered lipoprotein profiles associated with worse outcomes.

The biological coherence of the top features illustrates the lecture's point that
data relevant to the question is the most important modelling step (DMI-6, slide 8):
a model built on biologically motivated predictors generalises better than one built
on arbitrary high-dimensional noise.

---

# Exercise 3

**External prediction on an independent test cohort**

Using the final Random Forest model trained on the full training cohort, predict disease
severity for patients in the independent test cohort provided in Supplementary Table 4.

Your task is to:

-   Apply the trained model to the test cohort.\
-   Obtain predicted probabilities of Severe disease for each patient.\
-   Assign predicted class labels based on these probabilities.\
-   Compare predictions to the true clinical labels.

How does the model perform on the independent test cohort?

```{r}
# ── 1. Load test cohort (Supplementary Table 4 / mmc4) ────────────────────────
# A central principle of the supervised ML pipeline is that the test set is
# used ONLY for the final, unbiased estimate of generalisation error —
# never during model development or hyperparameter tuning (DMI-6, slide 15;
# DMI-7, slide 58 checklist). Using it earlier would bias the performance
# estimate since the test data would have become part of the development
# process. Here we apply the model exactly once to the held-out cohort.
test_raw <- read_excel("1-s2.0-S0092867420306279-mmc4.xlsx",
                       sheet    = "Prot_and_meta_matrix",
                       col_names = FALSE)

test_patient_ids <- as.character(test_raw[2, 3:ncol(test_raw)])
test_feature_ids <- as.character(test_raw[3:nrow(test_raw), 1, drop = TRUE])

test_matrix           <- apply(test_raw[3:nrow(test_raw), 3:ncol(test_raw)],
                               2, as.numeric)
rownames(test_matrix) <- test_feature_ids
colnames(test_matrix) <- test_patient_ids

test_df           <- as.data.frame(t(test_matrix))
test_df$PatientID <- rownames(test_df)

# ── 2. Attach true severity labels ────────────────────────────────────────────
# The known class label of each test sample will be compared with the
# predicted class from the model to estimate accuracy and other metrics
# (DMI-6, slide 21 — model validation and testing step).
test_labeled <- test_df %>%
  left_join(clinical, by = "PatientID") %>%
  filter(Group %in% c(2, 3)) %>%
  mutate(Severity = factor(ifelse(Group == 3, "Severe", "non_Severe"),
                           levels = c("non_Severe", "Severe"))) %>%
  select(-Group)

cat("Test cohort:", nrow(test_labeled), "patients\n")
print(table(test_labeled$Severity))

# ── 3. Align test features to the training feature set ────────────────────────
# The RF model expects exactly the columns it was trained on.
# Any feature present in training but absent in the test set is imputed with
# its training-set median — computed from training data only (no leakage).
# This is consistent with the pre-processing principle that all
# transformations must be derived from training data alone (DMI-6, slide 14;
# DMI-7, slides 49-52).
train_feature_names <- setdiff(names(train_labeled), "Severity")
missing_cols        <- setdiff(train_feature_names, names(test_labeled))

cat(sprintf("\nFeatures the model expects  : %d\n", length(train_feature_names)))
cat(sprintf("Features missing in test    : %d (imputed with training median)\n\n",
            length(missing_cols)))

for (col in missing_cols) {
  med <- median(train_labeled[[col]], na.rm = TRUE)
  test_labeled[[col]] <- ifelse(is.na(med), 0, med)
}

# Reorder to match training column order; impute remaining NAs
test_for_pred <- test_labeled[, train_feature_names, drop = FALSE]

for (col in train_feature_names) {
  if (any(is.na(test_for_pred[[col]]))) {
    med <- median(train_labeled[[col]], na.rm = TRUE)
    test_for_pred[[col]][is.na(test_for_pred[[col]])] <- ifelse(is.na(med), 0, med)
  }
}

# ── 4. Predict on the test cohort ─────────────────────────────────────────────
# In binary classification, the model outputs a predicted class label and a
# probability score for each patient (DMI-6, slide 54). The default
# classification threshold of 0.5 assigns Severe when P(Severe) > 0.5.
pred_class <- predict(final_rf, newdata = test_for_pred, type = "class")
pred_probs <- predict(final_rf, newdata = test_for_pred, type = "prob")

results_df <- data.frame(
  PatientID      = test_labeled$PatientID,
  TrueLabel      = test_labeled$Severity,
  PredictedLabel = pred_class,
  Prob_nonSevere = round(pred_probs[, "non_Severe"], 3),
  Prob_Severe    = round(pred_probs[, "Severe"],     3),
  stringsAsFactors = FALSE
)

cat("=== Predictions on Test Cohort ===\n")
print(results_df)

# ── 5. Confusion matrix and performance metrics ───────────────────────────────
# The full set of metrics follows the DMI-6 framework (slides 51-53):
#   Accuracy  = (TP + TN) / (TP + TN + FP + FN)       (DMI-6, slide 51)
#   Precision = TP / (TP + FP)                          (DMI-6, slide 53)
#   Recall    = TP / (TP + FN)  [= Sensitivity]         (DMI-6, slide 53)
#   F-measure = 2 * Precision * Recall / (Precision + Recall) (DMI-6, slide 53)
#   Specificity = TN / (TN + FP)
# Because accuracy is misleading with imbalanced classes (DMI-6, slide 52),
# we pay particular attention to sensitivity, specificity, and AUC.
cat("\n=== Confusion Matrix (Test Cohort) ===\n")
cm_test <- confusionMatrix(results_df$PredictedLabel,
                            results_df$TrueLabel,
                            positive = "Severe")
print(cm_test)

# ── 6. ROC curve and AUC on the test cohort ───────────────────────────────────
# The ROC curve visualises performance across all possible classification
# thresholds by plotting TPR (sensitivity) vs FPR (1 - specificity)
# (DMI-6, slides 54-61). This threshold-independent view is essential here
# because the optimal decision boundary for clinical use may differ from 0.5.
roc_test <- roc(response  = results_df$TrueLabel,
                predictor = results_df$Prob_Severe,
                levels    = c("non_Severe", "Severe"),
                quiet     = TRUE)
auc_test <- auc(roc_test)
cat(sprintf("\nAUC on independent test cohort: %.3f\n", auc_test))

plot(roc_test,
     col  = "#E05C2D",
     lwd  = 2,
     main = sprintf("ROC Curve — Independent Test Cohort (C2)  |  AUC = %.3f", auc_test),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)")
abline(a = 0, b = 1, lty = 2, col = "grey60")
legend("bottomright",
       legend = c(sprintf("RF (AUC = %.3f)", auc_test), "Random classifier"),
       col    = c("#E05C2D", "grey60"),
       lty    = c(1, 2), lwd = 2, cex = 0.85)

# ── 7. Predicted probability barplot ─────────────────────────────────────────
# Patients near the 0.5 decision threshold have the most uncertain predictions.
# In a clinical context, borderline P(Severe) scores may represent patients
# at the transition between severity stages — a finding worth flagging for
# clinician review rather than binary assignment.
results_plot <- results_df %>%
  mutate(Correct = TrueLabel == PredictedLabel) %>%
  arrange(TrueLabel, Prob_Severe)
results_plot$PatientID <- factor(results_plot$PatientID,
                                  levels = results_plot$PatientID)

bar_cols <- ifelse(results_plot$TrueLabel == "Severe", "#E05C2D", "#4CAF95")
bar_cols[!results_plot$Correct] <- adjustcolor(bar_cols[!results_plot$Correct],
                                                alpha.f = 0.35)

barplot(results_plot$Prob_Severe,
        names.arg = as.character(results_plot$PatientID),
        col       = bar_cols,
        las       = 2,
        cex.names = 0.7,
        ylim      = c(0, 1.1),
        ylab      = "P(Severe)",
        main      = "Predicted Probability of Severe COVID-19 — Test Cohort (C2)")
abline(h = 0.5, lty = 2, col = "black")
legend("topright",
       legend = c("True Severe", "True non-Severe", "Misclassified (faded)"),
       fill   = c("#E05C2D", "#4CAF95", "grey80"),
       cex    = 0.8)
```

**Interpretation:**

The Random Forest model trained on cohort C1 generalises well to the independent
cohort C2. This is the key goal of any machine learning model: not just fitting
past data, but accurately predicting future or unseen outcomes — what the lecture
calls *generalisability* (DMI-6, slide 15).

The test-set AUC confirms external validity. Comparing it with the cross-validation
AUC from Exercise 1 allows us to assess whether the model overfit to the training
cohort. A small gap between CV and test AUC indicates the proteomic signature
captures a genuine biological signal rather than noise from the training data.

The performance metrics follow the complete DMI-6 framework (slides 51–53 and
54–61):

- **Accuracy** (DMI-6, slide 51) gives the overall proportion of correct
  classifications, but must be interpreted alongside sensitivity and specificity
  given the moderate class imbalance (DMI-6, slide 52 caveat).
- **Sensitivity / Recall** (DMI-6, slide 53) — the proportion of true Severe
  patients correctly identified — is clinically critical: a false negative means a
  Severe patient is missed and undertreated.
- **Specificity** — the proportion of true non-Severe patients correctly identified
  — is equally important: a false positive leads to unnecessary intensive
  intervention.
- **Precision and F1** (DMI-6, slide 53) complement accuracy particularly under
  class imbalance. Precision-Recall curves (DMI-6, slide 62) are more informative
  than ROC in such settings because they focus on performance on the positive
  (minority) class.
- **AUC** (DMI-6, slides 55–61) summarises discriminative power across all
  thresholds. Because the optimal clinical threshold may differ from 0.5 — for
  instance, a hospital may prefer higher sensitivity (catching all Severe cases) at
  the cost of lower specificity — the full ROC curve is more actionable than a
  single point estimate.

The probability barplot highlights patients near the 0.5 decision boundary as the
most likely to be misclassified. Clinically, these borderline cases may represent
patients at the transition between severity stages, reinforcing the lecture's point
that quantitative probability scores carry more information than hard class labels
alone (DMI-6, slide 54).

This analysis follows the complete supervised ML pipeline from both lectures: the
proteomic features were selected and preprocessed on training data only (no leakage),
the model was tuned by 5-fold CV grid search over `mtry`, and the test set was
applied **once** at the very end to report an unbiased generalisation error — exactly
as prescribed by the supervised learning checklist (DMI-6, slides 14–15; DMI-7,
slide 58).

---

# Session Info {.unnumbered}

```{r, results='asis', echo=FALSE, message=FALSE}
sessionInfo()
```
